### 2020-09-08 paper/paper.tex
refactor paper

In many countries, particularly in the developing world, the quality of public education remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations. How can these educational outcomes be improved? Previous studies have found that incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on quasi-experimental research designs, these studies provide compelling evidence that actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers these public services and to whom do they respond?

@@

% # Related literature and debate

% In this section I review extant literature on public goods provision and the politics underlying it, focusing on more recent studies of bureaucratic personnel and political leaders reshape these institutions. I also highlight how my research incorporates multiplicity in political actors and how this affects bargaining over public sector jobs. I address this gap by adapting previous analyses of executive-legislative bargaining to bureaucratic control at the local level.

% ## Bureaucratic personnel and public goods provision

% Bureaucracies have a clear impact on the delivery of public services. A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions (centeno_unpacking_2017, kohli_state-directed_2004). A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy (johnson_miti_1982, kohli_state-directed_2004). A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth (evans_bureaucracy_1999).

% Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference. Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores. gulzar_politicians_2017 show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment. akhtari_political_2015, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.

% This recent wave of studies shed light on the intersection between politicians and bureaucracies. However, few of these studies explicitly model the multiple actors involved in managing bureaucracies. Understanding their diverse goals and action space provides a firm theoretical foundation to how different politicians can reshape bureaucracies. To do so I turn to the well-established literature on executive-legislative bargaining, applying its insights to the analysis of local government and administration.

% ## Presidential coalitionism and patronage

% For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors (grindle_jobs_2012, mccarty_appointments_2004). Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. (raile_executive_2011, power_optimism_2010).

% In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. (raile_executive_2011). In a setting characterized by weak party cohesion and programmatic commitments (ames_electoral_1995, lucas_ideological_2010), bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda (batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999).

% \blue{something about coalitions}

% In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies. Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.^[Interview with C, August and September 2019.] Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}. As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."^[Interview with C, August 2019.] These hires induce changes in personnel, with important consequences for bureaucracies and educational services.

% ## Bureaucratic turnover and inexperienced education

% Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects. As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population (gailmard_slackers_2007). Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher (clotfelter_teacher_2007.) akhtari_political_2015 finds that students attending a school with a recently appointed school principal perform worse in standardized test scores. When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones (colonnelli_patronage_2017).

% > "I am aware that the position is temporary. Especially because it is a political position, decided by the administration. If the current administration is out of power, we are automatically dismissed." - Interview with school principal A, August 2019.

% Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level. In this study, the primary focus is on bureaucrats working within the boundaries of a school: school principals and teachers. In the following section I describe the institutional context for public education in Brazil, as well as the data employed for the the estimation.

% # The argument

% ## The executive dilemma

% I model the executive decision over patronage as a product of political career concerns and strategic interaction with the local legislative power. In particular, mayors are assumed to be career-oriented and re-election seeking (besley_principled_2006). This provides an alternative set-up to extant treatments of patronage as a pre-electoral bargain, naturally highlighting the question of government performance (stokes_brokers_2013, robinson_political_2013.) The key feature of the model is the trade-off between patronage and quality of public services, which mayors are aware of when allocating public sector jobs to city councilors.

% The mayor's strategy centers around the redistribution public sector jobs. When the mayor takes office, a set of bureaucratic positions become available for allocation to city councilors. In a municipal parallel to presidential coalitionism, jobs are allocated to legislators in order to buy their vote. I assume, for now, that job allocations have a negative effect on public services. The mayor can refrain from distributing jobs for vote buying, since reducing turnover helps retain expertise and improves the quality of public services. However, doing so raises the threat of defection by coalition members, which can lead to the withholding of budgetary approval. This is the executive's dilemma.

% The main proposition is that mayors who control a smaller (larger) share of legislative seats allocate more (less) public sector job to city councilors. The intuition is as follows. Mayors need to secure a simple majority of legislative votes to have the budget approved. groseclose_buying_1996 and banks_buying_2000 point out that the executive should prioritize votes from cheaper legislators, which in this set-up are those who share the same electoral coalition as the mayor. Having exhausted these votes, the executive has to resort to buying votes from the opposition, which demand more patronage. The less seats occupied by members of the mayor's electoral coalition, the more expensive vote-buying becomes, and the observable implication is increased patronage in the municipality.

% The model builds on rich theoretical treatment of electoral accountability and majority-buying in legislatures. Starting with barro_control_1973 and ferejohn_incumbent_1986, political economists have proposed formal models analyzing the nexus of electoral accountability as an incentive for shaping the behavior of elected politicians (coate_form_1995; besley_principled_2006). The possibility of reelection can motivate them to exert effort and convince voters to reelect them. This model extends previous analyses by incorporating legislators to the game, with the need to secure a legislative majority to implement the executive agenda (groseclose_buying_1996, banks_buying_2000).

% ## The model

% There is one mayor $m$, a city council $N$ and representative median voter $v$. The goal of the mayor is to be reelected into office, yet for any given time period $m$ needs to garner legislative support. The city council plays that key role by approving or not the budget for the executive. The legislative chamber is composed by $N = {1,..,n}$ legislators, where $S \subseteq N$ denote the subset of legislators who belong to the electoral coalition of mayor $m$. For each $i \in N$, let $\textbf{v} = \{v_1,...,v_n\}$ describe legislator i's preference to support the mayor in the budgetary process. Let $v_i > v_j$ for all $i \in S$ and $j \in \{N \setminus S\}$. Therefore, members of the mayor's electoral coalition are more likely to approve the budget than outside members.

% I introduce a set of simplifying assumptions. First, mayors are one of two types: a technocratic politician who delivers high quality public goods, and a pragmatic one, who fails to do so. Note that this is an informational heuristic. Voters have beliefs about how competent a politician is and strictly prefer them over the alternative. Because voters cannot observe the type, they update beliefs observing the quality of public goods they observe (coate_form_1995). Additionally, I assume that the voter is sincere, excluding pivotality concerns and strategic voting from voters' utility calculations. Finally, I normalize the politician's exit option to 0: the politician always prefers to be re-elected.

% The game has two periods. Voter $v$ does not observe the amount of patronage handed out by mayor $m$ nor her type, but observes and derives utility from the realization of public good quality $\omega_t$. An important feature in the model is that high quality public services $\omega_{h}$ are produced probabilistically as a function of $\theta$, the quality of the bureaucracy. The more jobs a mayor allocates to legislators, the lower is $\theta$, thus capturing the inverse relation between patronage and state capacity.^[See kohli_state-directed_2004.] The budgetary process also affects the public service delivery, with a failure to approve the budget lowering the resources available to the executive.

% The mayor's optimal strategy is pinned down by the share of patronage which maximizes expected utility from holding onto office, conditional on her type and expected benefits from reelection. Mayor $m$ has to offer public sector jobs to coopt the opposition while ensuring that the voter is satisfied enough to reelect her. The executive dilemma is how to reconcile the incentive for reelection and the pressure imposed by the opposition.

% The timing of the game is as follows.

% 1) Nature draws the politician's type.
% 2) Politician observes her type and allocates public sector jobs to legislators, setting $\theta_1$.
% 4) Nature realizes public good $\omega_1(\theta_1)$.
% 5) Voter observes $\omega_1$ and casts vote to retain or fire the incumbent.
% 6) If incumbent is retained, she sets $\theta_2$ and nature draws $\omega_2(\theta_2)$. If incumbent is deposed, a challenger takes office, sets $\theta_2$ and the game ends.

% ## The executive dilemma

% The mayor's choice set is the bureaucratic quality parameter $\theta_t$. As quality increases - the mayor allocates a smaller proportion of jobs to legislators - the less votes become secure in the legislature. This reduces the likelihood of securing the necessary resources to provide public goods for a given fiscal year. Adapting gailmard_slackers_2007, there are two types of mayors: technocratic and pragmatic. Technocratic politicians invariably deliver high quality public goods to the voter. Pragmatic politicians, on the other hand, are susceptible to legislative pressure and at times fail to deliver them.

% The probability of a politician being technocratic is common knowledge: $\xi \in [0,1]$. Conversely, the probability of her being pragmatic is simply $1 - \xi$. Both types derive utility from public goods provision $\omega$, where for simplicity $u(\omega_l)=0$ and $u(\omega_h) = 1$. We do not model the decision of the competent mayor: she simply delivers high quality goods and is inevitably reelected. We focus on the more interesting case of the pragmatic mayor.

% \begin{align*}
% Eu_m \left(\theta_t, s_1 \middle|\lambda \right) &= \textbf{E}(\omega_1) + \delta \text{Pr}(\text{reelection})\left[\psi + \textbf{E}(\omega_2)  \right] \\
% &= \frac{1-s_1\theta_1}{1+e^{-(10\theta_1-5)}} + \delta \text{Pr}(\text{reelection})\left[\psi + \frac{1-s_2\theta_2}{1+e^{-(10\theta_2-5)}} \right] \\
% \end{align*}

% The equation above formalizes the mayor's dilemma. The probability of reelection decreases with the share of jobs allocated to the opposition, since it lowers the quality of the bureaucracy. However, if the mayor fails to appease them through patronage, it is increasingly costly to "govern". While technocrats are immune to these political pressures, pragmatists ultimately cave in. The need to coopt legislators decreases the quality of the bureaucracy and public goods, as mayors are forced to allocate positions within the bureaucracy to garner support in the legislative. Details on deriving the equilibrium strategy are in the appendix. I now describe Brazil as a case for testing the theory.

% # Institutional context and data

% ## Education in Brazil

% In Brazil, responsibility for primary education is delegated to local governments (paschoal_historia_2009). The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments. Figure \ref{fig:enrollment} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

% \blue{plots/enrollment_dep.pdf}

% The local executive branch has sole prerogative over the hiring and firing of teachers and school principals. School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (brollo_political_2017). Teachers usually enter the public service through a public exam and are eligible for tenure after two years (gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount. Municipal teachers and school principals are overseen by a local department of education (militao_educacao_2014).

% Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education. At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (ferraz_corrupting_2012).

% ## Local governance and mayoral coalitions

% Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council. Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.^[The last municipal elections were in 2016. These take place ] The executive is responsible for the implementation of social policies such as education, as well as drafting the municipal budget. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year. As noted by souza_reforma_2004, decisions over how to manage education are primarily in the hands of the mayor, but the city councilors exert significant pressure.

% In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (dantas_eleicoes_2013). These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber. Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}). While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.^[Interview with C, former chief of staff of municipality J. September 2018.]

% Interviews with secretaries of education and mayors suggest that city councilors play an important role in staffing decisions. While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves. These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council. In the words of a set of school administrators in the municipality of I:

% > School principal: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.

% > Educational counselor: I was also invited to work here, by the city councilor and the department of education.^[Interview with the administrative board of school A., municipality J. August 2019]

% These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff. To verify these claims in a broader set of municipalities, as well as outlining the research design, I employ extensive administrative and electoral data collected by the Brazilian federal government. In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.

>>>>

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations. How can these educational outcomes be improved? Previous studies have found that incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on quasi-experimental research designs, these studies provide compelling evidence that actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers these public services and to whom do they respond?

@@

\section{Related literature}

In this section I review extant literature on public goods provision and the politics underlying it, focusing on more recent studies of bureaucratic personnel and political leaders reshape these institutions. I also highlight how my research incorporates multiplicity in political actors and how this affects bargaining over public sector jobs. I address this gap by adapting previous analyses of executive-legislative bargaining to bureaucratic control at the local level.

\subsection*{Bureaucratic personnel and public goods provision}

Bureaucracies have a clear impact on the delivery of public services. A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions \citep{centeno_unpacking_2017, kohli_state-directed_2004}. A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy \citep{johnson_miti_1982, kohli_state-directed_2004}. A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth \citep{evans_bureaucracy_1999}.

Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference. Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores. \citet{gulzar_politicians_2017} show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment. \citet{akhtari_political_2015}, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.

This recent wave of studies shed light on the intersection between politicians and bureaucracies. However, few of these studies explicitly model the multiple actors involved in managing bureaucracies. Understanding their diverse goals and action space provides a firm theoretical foundation to how different politicians can reshape bureaucracies. To do so I turn to the well-established literature on executive-legislative bargaining, applying its insights to the analysis of local government and administration.

\subsection*{Presidential coalitionism and patronage}

For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle_jobs_2012, mccarty_appointments_2004}. Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. \citet{raile_executive_2011, power_optimism_2010}.

In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. \citep{raile_executive_2011}. In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.

\blue{something about coalitions}

In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies. Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}. As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.

\subsection*{Bureaucratic turnover and inexperienced education}

Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects. As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population \citet{gailmard_slackers_2007}. Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher \citet{clotfelter_teacher_2007}. \citet{akhtari_political_2015} finds that students attending a school with a recently appointed school principal perform worse in standardized test scores. When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones \citet{colonnelli_patronage_2017}.

> "I am aware that the position is temporary. Especially because it is a political position, decided by the administration. If the current administration is out of power, we are automatically dismissed." - Interview with school principal A, August 2019.

Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level. In this study, the primary focus is on bureaucrats working within the boundaries of a school: school principals and teachers. In the following section I describe the institutional context for public education in Brazil, as well as the data employed for the the estimation.

\section{Institutional context and data}

\subsection{Education in Brazil}

In Brazil, responsibility for primary education is delegated to local governments (paschoal_historia_2009). The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments. Figure \ref{fig:enrollment} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

\blue{plots/enrollment_dep.pdf}

The local executive branch has sole prerogative over the hiring and firing of teachers and school principals. School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (brollo_political_2017). Teachers usually enter the public service through a public exam and are eligible for tenure after two years (gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount. Municipal teachers and school principals are overseen by a local department of education (militao_educacao_2014).

Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education. At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (ferraz_corrupting_2012).

\subsection{Local governance and mayoral coalitions}

Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council. Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.^[The last municipal elections were in 2016. These take place ] The executive is responsible for the implementation of social policies such as education, as well as drafting the municipal budget. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year. As noted by souza_reforma_2004, decisions over how to manage education are primarily in the hands of the mayor, but the city councilors exert significant pressure.

In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (dantas_eleicoes_2013). These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber. Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}). While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.^[Interview with C, former chief of staff of municipality J. September 2018.]

Interviews with secretaries of education and mayors suggest that city councilors play an important role in staffing decisions. While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves. These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council. In the words of a set of school administrators in the municipality of I:

> School principal: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.

> Educational counselor: I was also invited to work here, by the city councilor and the department of education.^[Interview with the administrative board of school A., municipality J. August 2019]

These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff. To verify these claims in a broader set of municipalities, as well as outlining the research design, I employ extensive administrative and electoral data collected by the Brazilian federal government. In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.

\section{Theory}

In this section, I apply a canonical vote-buying model in legislatures \citep{groseclose_1996_buying, banks_2000_buying} to model mayor's strategy to use patronage to buy legislative votes. For any given electoral term, a mayor wants to enact her preferred policy, but to do so needs to pass it through a simple majority in the city council. To ensure victory, the mayor offers appointment positions to voters, in this case, city councilors, who in exchange vote in favor of the bill.

In this setting, passing a policy proposal requires outbidding the opposition, which generates interesting results regarding the optimal size of the coalition and conditions under which we should observe minimal, super and universal coalitions. In particular, the amount of resources that the opposition controls can force the mayor's hands and increase the amount of patronage necessary to pass the bill. We pursue these empirical implications of the model further in our results section.

The amount of public sector positions the mayor offers to these city councilors also responds to the composition of the city council. The intuition is straightforward: if the city councilors are already likely to support the bill, there is no need to offer them additional side payments. Note that there is no claim about a dichotomy of clientelism, in which there is or is not patronage \citep{stokes_brokers_2013, grindle_jobs_2012}, but rather a variation in degrees: how many appointments must the mayor give out in order to govern on her preferred policy.

For the remainder of this section, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper. I conclude with a discussion of the model and how it applies to the Brazilian municipal context.

\subsection*{The Setting}

The government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.\footnote{Note that I use the terms \emph{mayor} and \emph{government} interchangeably.} There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition. In order to implement her policy the mayor must pass a simple majority vote in the city council, comprised of an odd number $N$ of legislators. The total amount of political resources available is $W_G$ and $W_O$, which for the mayor includes public sector appointments.

Each city councilor is characterized by a publicly observed policy preference $v_i$ for all $i \in N$, where $v_i > 0$ denotes that the mayor's proposal $x$ is preferred by legislator $i$. Let $\mathbf{v} = (v_1, ..., v_n)$ describe the preference profile for the city councilors. In this setting, $v_i$ measures the degree to which an individual city councilor supports the mayor, with higher values of $v$ denoting stronger support for the mayor and vice versa. Payoff are realized when city councilor $i$ votes, independent of the outcome of the voting procedure. This sincere voting preference closes the possibility of general equilibria in which $i$'s voting behavior affect $j$.

We solve the game through backward induction. The timing of the game is as follows:

\begin{enumerate}
    \item Government $G$ offers a bribe schedule $a \in (a_1, ..., a_n) \in \real$.
    \item Opposition $O$ observes the bribe schedule $m$ and makes a counter-offer $b \in (b_1, ..., b_n) \in \real$.
    \item City councilors cast their votes at the end of bribing period.
    \item Nature sums legislative votes, legislative outcome is decided and payoffs are realized.
\end{enumerate}

Given a bribe schedule $(a, b)$, councilor $i$ prefers to vote for the mayor's proposal $x$ if $a_i + v_i > b_i$ and the status quo $y$ otherwise. Since indifferent councilors vote for the status quo, the opposition needs to only match bribes from $M$, adjusting for individual preferences, i.e. $b_i = a_i + v_i$. For the mayor, she needs to construct the cheapest winning coalition in order to defeat the opposition.

Following Groseclose and Snyder (1996) and Banks (2000) we focus our analysis on the set of equilibria in which the mayor wins.\footnote{Since strategies for both players are symmetrical, any set of equilibria in which the mayor loses can be modeled as cases in which the the opposition loses.} In this context, the amount of patronage resources $W_G$ is sufficiently large relative to $W_O$ and $\bfv$ that the mayor's preferred policy $x$ is implemented over $y$. Let $U(\bfv, W_O)$ denote the set of unbeatable patronage schedules for the mayor, and for any patronage schedule let $S(a) = \sum_{i = 1}^n a_i$ be the total amount of patronage disbursed. The mayor then solves

\begin{equation}
    \label{eqn:solution}
    \min_a\{S(a) : a \in U(\mathbf{v}, W_B) \}
\end{equation}

Note that for any equilibrium strategy, it must be the case that mayor $M$ uses a leveling schedule: every city councilor in her coalition $C$ is equally expensive for the opposition $O$ to bribe. More formally, for any $a \in \mathbb{R}_+^n$, let $C(a) = \{i \in N : a_i > 0\}$ denote the set of individuals who receive a bribe from from the government $G$. One can show that there is a bribe schedule $a'$ such that for any $i,j \in C(a)$, $a'_i + v_i = a'_j + v_j$. The intuition is that the mayor has no incentive to make voters differentially expensive, because the opposition $O$ will simply ignore the more expensive voters and target the least favorable members of the coalition. We refer to these strategies as leveling schedules.

We can characterize the set of equilibria in the game by introducing additional notation. Let $U^l(\mathbf{v}, W_O) \subseteq U(\mathbf{v}, W_O)$ denote the set of unbeatable leveling schedules. These are bribe schedules such that $a_i + v_i = a_j + v_j  \equiv t(a)$. The bribe $a_i = t(a) - v_i$ is the sum of two terms. The first is the common "transfer" among all voters in $C(a)$, the second ($-v_i$) is individual specific. The latter term makes voters indifferent between $x$ and $y$ absent any bribe from $B$; the former represents the per capita amount necessary to make $C(a)$, together with any unbribed voters, unaffordable for $B$.

I impose the following two assumptions:
\begin{align*}
    A1 &: v_{(n+1)/2} < 0\\
    A2 &: v_1 < 2W_B/(n+1) 
\end{align*}
$A1$ implies that absent any bribes by $A$, $y$ will defeat $x$. Therefore $A$ must bribe at least one voter. $A2$ implies that $A$ must bribe at least a majority of voters, otherwise $B$ will have sufficient resources to bribe $(n+1)/2$ voters and win. $A2$ also implies that for all $a \in U^l(\mathbf{v}, W_B)$, it must be that $t(a) \geq 2W_B/(n + 1)$, otherwise $B$ can bribe a majority from $C(a)$ itself and win the vote.

% For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$. Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$. Then, under $A2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$: $a'_i = t(a) - v_i$, $a'_j = 0$ and $a'_m = a_m$ for all $m \notin i, j$.\footnote{Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e. $a'_i \leq a_j$. Also, because of $A2$ $a'_j \leq a_j$, since it guarantees $a_j$ is non-negative.} Generalizing, and recalling that $v_1 \geq ... \geq v_n$, we see that for all $a \in U^l(\mathbf{v}, W_b)$ there exists a bribe schedule $a' \in U^l(\mathbf{v}, W_b)$ such that $S(a') \leq S(a)$ and $C(a') = \{1, ..., k(a)\}$. Therefore, we can without loss of generality restrict attention to schedules $a$ by $A$ which bribe the first $k(a)$ voters. Call these monotonic leveling schedules and let $U_m^l \subseteq U(\mathbf{v}, W_B)$.

These assumptions allow us to restrict our analysis to unbeatable monotonic leveling schedules, which we denote as $U_m^l$.\footnote{A detailed explanation can be found in the appendix.} We can simplify the total expenditure on patronage by the government, $S(a)$, as

\begin{equation*}
    S(a) = \sum_{i \in C(a)}a_i = k(a) \cdot t(a) - \sum_{i \leq k(a)}v_i
\end{equation*}

Note that the choice of $k(a)$ and $t(a)$ fully characterize any schedule $a \in U_m^l(\mathbf{v}, W_B)$. We can thus fully characterize the optimization problem of $A$ in equation \ref{eqn:solution} as

\begin{equation*}
    \min_{k,t} k \cdot t- \sum_{i \leq k} v_i
\end{equation*}

subject to the constraint that the induced schedule $a \in U_m^l$. We can reformulate this as an unconstrained problem by noting the following. First, if $a(k, t, \mathbf{v})$ is unbeatable, it must be that $k \geq (n + 1)/2$, so by $A1$ it must be that if $a_i(k, t, \mathbf{v}) = 0$, then $v_i < 0$. Therefore, $B$ receives all non-bribed voters for free. For $a(k, t, \mathbf{v})$ to be unbeatable, then, it must be that $B$ cannot afford the remaining $(n + 1)/2 - (n - k) = k - (n - 1)/2$ voters, or

$$t \cdot (k - (n - 1)/2) \geq W_B$$

Solving this for equality yields the optimal transfer from $A$ to members of $C(A) = \{1, ..., k\}$, conditional on k:

\begin{align}
    \label{eqn:common_transfer}
    t(k, W_B) = \frac{W_B}{k - (n - 1)/2}
\end{align}

Defining minimal winning expenditures as

\begin{align}
    \label{eqn:total_expenditure}
    E(k, \mathbf{v}, W_B) = k \cdot t(k, W_B) - \sum_{i \leq k} v_i
\end{align}

we can state $A$'s problem as

\begin{align}
\label{eqn:minimal_expenditure}
\min_k \{E(k, \mathbf{v}, W_B) : k \in {(n + 1/2), ..., n}\}
\end{align}

Denote the solution to expression \ref{eqn:minimal_expenditure} as $k^*(\mathbf{v}, W_B)$. This solution implicitly generates a solution to expression \ref{eqn:solution} through expression \ref{eqn:total_expenditure} and the induced bribe schedule above. Therefore, characterizing the optimal $k^*$ is sufficient to fully characterize the optimal behavior of the mayor.

\subsection*{Results}

First, characterize a solution for $k^*$. Because $k$ is finite, calculus cannot be employed. Instead, we deploy a discrete version of these techniques. First let's define $\Delta(k) = E(k + 1) - E(k)$ as the difference in expenditure from adding another coalition member. Note that if $\Delta(k) \geq 0$ then $A$ does not want to add another member to the coalition. Conversely, if $\Delta(k) < 0$, then $A$ is strictly better off by adding the $k + 1$th member of the coalition.

Next, suppose that $\Delta(k)$ is increasing in $k$. The following algorithm can then be used to identify $k^*$: if $\Delta((n + 1)/2) \geq 0$, then we know from $\Delta(k)$ increasing that $A$ is better off by setting $k^*$ to $(n + 1)/2$. If $\Delta((n + 1)/2) < 0$, then we know that $k^*$ must be greater than $(n + 1)/2$, so we next solve for $\Delta((n + 3)/2)$, and so on.

We can therefore search for the optimal $k^*$ with the following algorithm:

\begin{align}
    \label{eqn:optimal_k}
    k^* = 
    \begin{cases}
        (n + 1)/2 & \text{if } \Delta((n + 1)/2) \geq 0 \\
        \max\{k : \Delta(k - 1) < 0\} & \text{otherwise}
    \end{cases}
\end{align}

We can also further characterize the change in minimum winning expenditures in equation \ref{eqn:total_expenditure} as

\begin{align}
    \Delta(k)  & = \left[\frac{(k + 1)W_B}{k + 1 - (n - 1)/2} - \sum_{i \leq k + 1}v_i \right]\\
    \label{eqn:deltAk}
    & = \frac{-W_B (n - 1)}{2(k + 1 - (n - 1)/2)(k - (n - 1)/2)} - v_{k + 1}\\
    \label{eqn:def_deltAk}
    & \equiv T(k, W_B) - v_{k + 1}
\end{align}

Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.
